---
title: "Weather and Corn Yield Regressions"
author: "Bri Rick"
date: "3/3/2022"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
library(tidyverse)
library(R.matlab)
library(rnassqs)
library(dplyr)
```

## Weather Data Analysis

### Load the PRISM daily maximum temperatures

```{r tmax data}

# daily max temperature
# dimensions: counties x days x years
prism <- readMat("data/prismiowa.mat")

# look at county #1
t_1981_c1 <- prism$tmaxdaily.iowa[1,,1]
t_1981_c1[366]
plot(1:366, t_1981_c1, type = "l")

ggplot() +
  geom_line(mapping = aes(x=1:366, y = t_1981_c1)) +
  theme_bw() +
  xlab("day of year") +
  ylab("daily maximum temperature (°C)") +
  ggtitle("Daily Maximum Temperature, Iowa County #1")


```
```{r tidying up}

# assign dimension names to tmax matrix
dimnames(prism$tmaxdaily.iowa) <- list(prism$COUNTYFP, 1:366, prism$years)

# converted 3d matrix into a data frame
tmaxdf <- as.data.frame.table(prism$tmaxdaily.iowa)

# relabel the columns
colnames(tmaxdf) <- c("countyfp","doy","year","tmax")
tmaxdf <- tibble(tmaxdf)

```

## Temperature trends

### Summer temperature trends: Winneshiek County

```{r temp trends}

tmaxdf$doy <- as.numeric(tmaxdf$doy)
tmaxdf$year <- as.numeric(as.character(tmaxdf$year))

winnesummer <- tmaxdf %>%
  filter(countyfp==191 & doy >= 152 & doy <= 243) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

ggplot(winnesummer, mapping = aes(x = year, y = meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "Tmax (°C)") +
  geom_smooth(method = lm)

lm_summertmax <- lm(meantmax ~ year, winnesummer)
summary(lm_summertmax)

```

### Winter Temperatures - Winneshiek County

```{r winter temps}

winnewinter <- tmaxdf %>%
  filter(countyfp==191 & (doy <= 59 | doy >= 335) & !is.na(tmax)) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

ggplot(winnewinter, mapping = aes(x = year, y = meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "Tmax (°C)") +
  geom_smooth(method = lm)

lm_wintertmax <- lm(meantmax ~ year, winnewinter)
summary(lm_wintertmax)

```

### Multiple regression -- Quadratic time trend

```{r quadratic temp trend}

winnewinter$yearsq <- winnewinter$year^2

lm_wintertmaxquad <- lm(meantmax ~ year + yearsq, winnewinter)
summary(lm_wintertmaxquad)
winnewinter$fitted <- lm_wintertmaxquad$fitted.values

ggplot(winnewinter) +
  geom_point(mapping = aes(x = year, y = meantmax)) +
  geom_line(mapping = aes(x = year, y = fitted)) +
  theme_bw() +
  labs(x = "year", y = "tmax")

```

### Download NASS corn yield data

```{r yield download}

# set our API key with NASS
nassqs_auth(key = "F7506B86-1C9E-39F3-AA20-719CAD3B47A3")

# parameters to query on 
params <- list(commodity_desc = "CORN", util_practice_desc = "GRAIN", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")

# download
cornyieldsall <- nassqs_yields(params)

cornyieldsall$county_ansi <- as.numeric(cornyieldsall$county_ansi)
cornyieldsall$yield <- as.numeric(cornyieldsall$Value)

# clean and filter this dataset
cornyields <- select(cornyieldsall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
cornyields <- tibble(cornyields)

```

## Assignment

### Question 1a: Extract Winneshiek County corn yields, fit a linear time trend, make a plot. Is there a significant time trend?

*Yes, there is a significant increase in corn yield over time for Winneshiek County, with an increase of about 2.46 per year. *

```{r}
cornWinn <- cornyields %>% filter(county_name == "WINNESHIEK")

lm_cornWinn <- lm(yield ~ year, cornWinn)
cornWinn$fitted <- lm_cornWinn$fitted.values
summary(lm_cornWinn)

ggplot(cornWinn) +
  geom_point(aes(x = year, y = yield)) +
  geom_line(aes(x = year, y = fitted)) +
  theme_bw()
```


### Question 1b: Fit a quadratic time trend (i.e., year + year^2) and make a plot. Is there evidence for slowing yield growth? 

*There does not appear to be evidence for a slowing in yield growth in Winnesheik County in Iowa*

```{r}
cornWinn$yearsq <- cornWinn$year^2

lm_cornWinnquad <- lm(yield ~ year + yearsq, cornWinn)
summary(lm_cornWinnquad)
cornWinn$fittedsq <- lm_cornWinnquad$fitted.values

ggplot(cornWinn) +
  geom_point(aes(x = year, y = yield)) +
  geom_line(aes(x = year, y = fittedsq)) +
  theme_bw()
```


### Question 2 -- Time Series: Let's analyze the relationship between temperature and yields for the Winneshiek County time series. Use data on yield and summer avg Tmax. Is adding year or Tmax^2 to your model helpful? Make a plot and interpret the results.

*Adding tmax^2 to the model is helpful and gives a better fit for the model.*

```{r}
tmaxyield <- cornWinn %>% inner_join(winnesummer, by = "year")
tmaxyield$tsq <- tmaxyield$meantmax^2

lm_tmaxyield <- lm(yield ~ meantmax, tmaxyield)
summary(lm_tmaxyield)
tmaxyield$fitmeantmax <- lm_tmaxyield$fitted.values

lm_tmaxyieldyear <- lm(yield ~ meantmax + year, tmaxyield)
summary(lm_tmaxyieldyear)
tmaxyield$fitmeantmaxyear <- lm_tmaxyieldyear$fitted.values

lm_tmaxyieldsq <- lm(yield ~ meantmax + tsq, tmaxyield)
summary(lm_tmaxyieldsq)
tmaxyield$fitmeantmaxyield <- lm_tmaxyieldsq$fitted.values

ggplot(tmaxyield) +
  geom_point(aes(y = yield, x = meantmax)) +
  geom_line(aes(x = meantmax, y = fitmeantmax))+
  theme_bw()
  
ggplot(tmaxyield) +
  geom_point(aes(y = yield, x = meantmax)) +
  geom_line(aes(x = meantmax, y = fitmeantmaxyear))+
  theme_bw()
  
ggplot(tmaxyield) +
  geom_point(aes(y = yield, x = meantmax)) +
  geom_line(aes(x = meantmax, y = fitmeantmaxyield))+
  theme_bw()
```


### Question 3 -- Cross-Section: Analyze the relationship between temperature and yield across all counties in 2018. Is there a relationship? Interpret the results.

*There is an increase in yield with temperature until about 15 degrees C, then there is a decrease in yield with an increase in temperature. A simple linear model does not fit this trend well, however, a quadradic model incorporating mean T max and mean T max squared fits well.*

```{r}
cornyields18 <- cornyields %>% filter(year == 2018) %>% 
  group_by(county_name)

tmax_county <- tmaxdf %>% group_by(countyfp) %>% 
  filter(year == 2018) %>% 
  filter(doy >- 152 & doy <= 243) %>% 
  summarize(meantmax = mean(tmax)) %>% 
  rename(county_ansi = "countyfp") 

tmax_county$county_ansi <- as.numeric(as.character(tmax_county$county_ansi))

yield_county <- left_join(cornyields18, tmax_county, by = "county_ansi")
yield_county$tsq <- yield_county$meantmax^2

lm_counties <- lm(yield ~ meantmax, yield_county)
summary(lm_counties)
yield_county$fitted <- lm_counties$fitted.values

lm_countiessq <- lm(yield ~ meantmax + tsq, yield_county)
summary(lm_tmaxyieldsq)
yield_county$fitmeantmaxyield <- lm_countiessq$fitted.values

ggplot(yield_county) +
  geom_point(aes(x=meantmax, y = yield)) +
  geom_line(aes(x = meantmax, y = fitted)) +
  geom_line(aes(x = meantmax, y = fitmeantmaxyield)) +
  xlab("Mean T max")+
  ylab("Corn yield")+
  theme_bw()

```


### Question 4 -- Panel: One way to leverage multiple time series is to group all data into what is called a "panel" regression. Convert the county ID code ("countyfp" or "county_ansi") into factor using as.factor, then include this variable in a regression using all counties' yield and summer temperature data. How does the significance of your temperature coefficients (Tmax, Tmax^2) change? Make a plot comparing actual and fitted yields and interpret the results of your model.

*The fitted yield tends to be higher than the actual yield for corn in Iowa.*

```{r}
cornyieldsall <- cornyields %>% group_by(county_ansi) 

tmaxcountyall <- tmaxdf %>%  group_by(countyfp) %>% 
  filter(doy >- 152 & doy <= 243) %>% 
  summarize(meantmax = mean(tmax)) %>% 
  rename(county_ansi = "countyfp")

tmaxcountyall$county_ansi <- as.numeric(as.character(tmaxcountyall$county_ansi))

panel <- left_join(cornyieldsall, tmaxcountyall, by = "county_ansi")
  
panel$county_ansi <- as.factor(panel$county_ansi)
panel$tsq <- panel$meantmax^2

lm_yield_county_combo <- lm(yield ~ meantmax + tsq + county_ansi + year, panel)
summary(lm_yield_county_combo)
panel$fitted <- lm_yield_county_combo$fitted.values

ggplot(panel)+
  geom_point(aes(x = yield, y = fitted)) +
  geom_smooth(aes(x = yield, y = fitted), method = lm) +
  xlab("Actual Yield")+
  ylab("Fitted Yield")+
  theme_bw()
```


### Question 5 -- Soybeans: Download NASS data on soybean yields and explore either a time series relationship for a given county, the cross-sectional relationship for a given year, or a panel across all counties and years.

*For soybeans, the fitted values for yields less than about 40 overestimate the acutal yields, whereas values greater than 40 tend to underestimate the acutal yield values.*

```{r}
# set our API key with NASS
nassqs_auth(key = "F7506B86-1C9E-39F3-AA20-719CAD3B47A3")

# parameters to query on 
paramssoy <- list(commodity_desc = "SOYBEANS", statisticcat_desc="YIELD", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")

# download
soyyieldsall <- nassqs_yields(paramssoy)

soyyieldsall$county_ansi <- as.numeric(soyyieldsall$county_ansi)
soyyieldsall$yield <- as.numeric(soyyieldsall$Value)

# clean and filter this dataset
soyyields <- select(soyyieldsall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
soyyields <- tibble(soyyields)

soyyields <- soyyieldsall %>% group_by(county_code) %>% 
  unique() %>% 
  filter(!is.na(county_ansi))

tmaxcountyall <- tmaxdf %>%  group_by(countyfp) %>% 
  filter(doy >- 152 & doy <= 243) %>% 
  summarize(meantmax = mean(tmax)) %>% 
  rename(county_ansi = "countyfp")

tmaxcountyall$county_ansi <- as.numeric(as.character(tmaxcountyall$county_ansi))

panelsoy <- left_join(soyyields, tmaxcountyall, by = "county_ansi")
  
panelsoy$county_ansi <- as.factor(panelsoy$county_ansi)
panelsoy$tsq <- panelsoy$meantmax^2

lm_yield_county_combo <- lm(yield ~ meantmax + tsq + county_ansi + year, panelsoy)
summary(lm_yield_county_combo)
panelsoy$fitted <- lm_yield_county_combo$fitted.values

ggplot(panelsoy)+
  geom_point(aes(x = yield, y = fitted)) +
  geom_smooth(aes(x = yield, y = fitted), method = lm) +
  xlab("Actual Yield")+
  ylab("Fitted Yield")+
  theme_bw()
```

